---
title: "Demonstration 2: scATAC-seq Simulation Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration 2: scATAC-seq Simulation Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Single-cell assay for transposase-accessible chromatin sequencing (scATAC-seq) measures chromatin accessibility at single-cell resolution. Unlike scRNA-seq expression counts, scATAC-seq data have unique properties:
* **High Sparsity:** Over 90% to 98% of values in peak-by-cell matrices are zeroes because each locus has at most 2 copies per cell.
* **Near-Binary Nature:** Values primarily represent open (accessible) vs. closed chromatin states.
* **Peak Co-Accessibility:** Genomic loci regulated together exhibit correlated accessibility across cells.

This tutorial shows how to evaluate synthetic scATAC-seq data using **`scSimEval`**.

---

# 1. Loading the Packaged scATAC-seq Dataset

`scSimEval` includes an empirical reference and simulated scATAC-seq peak accessibility dataset (`example_scatac`):

```{r setup}
library(scSimEval)

# Load packaged scATAC-seq example data
data(example_scatac)

cat("Reference scATAC-seq dimensions:", dim(example_scatac$ref), "\n")
cat("Simulated scATAC-seq dimensions:", dim(example_scatac$sim), "\n")
cat("Cell types:", levels(example_scatac$cell_types), "\n")
cat("Batches:", levels(example_scatac$batch_info), "\n")
```

The matrices contain integer insertion counts across genomic peaks ($60\text{ peaks} \times 80\text{ cells}$).

---

# 2. Epigenomic Distribution & Summary Properties

We evaluate whether simulated peak insertion frequencies and cell library sizes match real distributions using `evaluate_simulation_accuracy()`:

```{r dist-eval}
atac_res <- evaluate_simulation_accuracy(
  ref_data          = example_scatac$ref,
  sim_data          = example_scatac$sim,
  compute_bivariate = FALSE,
  verbose           = FALSE
)

# Inspect peak accessibility and library size statistical distances
head(atac_res$metrics_summary_table, 8)
```

---

# 3. Epigenomic Cluster & Cell-Type Concordance

In scATAC-seq, identifying cell types requires clustering cells based on chromatin accessibility profiles. We test whether the simulator preserves genuine biological separation using `evaluate_clustering_metrics()`:

```{r clust-eval}
atac_clust <- evaluate_clustering_metrics(
  data       = example_scatac$sim,
  cell_types = example_scatac$cell_types,
  ref_data   = example_scatac$ref
)

cat("Average Silhouette Width (Simulated):", round(atac_clust$silhouette, 4), "\n")
cat("Davies-Bouldin Index:", round(atac_clust$davies_bouldin, 4), "\n")
cat("Adjusted Rand Index (ARI):", round(atac_clust$ARI, 4), "\n")
cat("Normalized Mutual Information (NMI):", round(atac_clust$NMI, 4), "\n")
```

---

# 4. Epigenomic Batch Integration & Confounder Handling

When simulating multi-sample experiments, batch variations can confound biological signals. We evaluate batch mixing metrics using `evaluate_batch_metrics()`:

```{r batch-eval}
atac_batch <- evaluate_batch_metrics(
  data       = example_scatac$sim,
  batch_info = example_scatac$batch_info,
  cell_types = example_scatac$cell_types,
  verbose    = FALSE
)

cat("Batch Shannon Entropy:", round(atac_batch$shannon_entropy, 4), "\n")
cat("Principal Component Regression (PCR R2):", round(atac_batch$pcr_r2, 4), "\n")
cat("Cross-Batch Transfer Accuracy:", round(atac_batch$cross_batch_accuracy, 4), "\n")
```

---

# 5. Chromatin Peak Co-Accessibility & Regulatory Coupling

In real biological cells, peaks that are close to each other or regulated by the same transcription factor complexes show correlated accessibility:

```{r peak-coacc}
# Evaluate peak co-accessibility fidelity
coacc_res <- calc_peak_coaccessibility_fidelity(
  ref_atac = example_scatac$ref,
  sim_atac = example_scatac$sim
)

cat("Co-Accessibility RV Coefficient:", round(coacc_res$rv_coefficient, 4), "\n")
cat("Peak Co-Accessibility Pearson Correlation:", round(coacc_res$coaccessibility_pearson, 4), "\n")
cat("Peak Co-Accessibility Spearman Correlation:", round(coacc_res$coaccessibility_spearman, 4), "\n")
cat("Frobenius Matrix Distance:", round(coacc_res$frobenius_distance, 4), "\n")
```

---

# 6. Master Unimodal scATAC-seq Pipeline

Execute the comprehensive evaluation workflow in a single coordinated command:

```{r master-eval}
master_atac <- evaluate_simulation_accuracy(
  ref_data          = example_scatac$ref,
  sim_data          = example_scatac$sim,
  elapsed_time      = 18.2,
  memory_mb         = 512.4,
  compute_bivariate = FALSE,
  verbose           = FALSE
)

# Inspect tidy summary table
head(master_atac$metrics_summary_table)
```

---

# 7. Conclusion

`scSimEval` effectively benchmarks scATAC-seq simulations, validating peak insertion distributions, extreme sparsity, cell type clustering, batch mixing, and chromatin peak co-accessibility without requiring synthetic ground-truth annotations.
