---
title: "Demonstration 3: Multiomics (scRNA-seq + scATAC-seq) Simulation Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration 3: Multiomics (scRNA-seq + scATAC-seq) Simulation Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Single-cell multiomics technologies allow simultaneous measurement of multiple cellular features, such as gene expression (scRNA-seq) and chromatin accessibility (scATAC-seq). Simulating multiomics data presents unique challenges because algorithms must preserve not only the distinct statistical properties of each individual modality, but also the biological coupling and regulatory connections between them.

---

# Paired vs. Unpaired Multiomics: How `scSimEval` Works

A common question is whether `scSimEval` supports both **paired** and **unpaired** single-cell multiomics data. 

**The answer is yes — `scSimEval` provides comprehensive evaluation for both data types:**

### 1. Paired Multiomics (e.g., 10x Chromium Multiome, SHARE-seq, SNARE-seq)
* **Definition:** Both scRNA-seq expression counts and scATAC-seq chromatin accessibility peaks are measured simultaneously from the **exact same individual cells** (each column in the RNA matrix corresponds to the same cell barcode in the ATAC matrix).
* **Evaluation in `scSimEval`:**
  * Runs all unimodal evaluations across both modalities (Categories 1 through 6 and Category 8).
  * **Unlocks full Category 7 (Cross-Modal Coupling):**
    * **FOSCTTM (Fraction of Samples Closer Than The True Match):** Evaluates whether the paired cell is its nearest neighbor in joint embedding space.
    * **Match@1 Rate:** Percentage of cells whose true multiomics profile is rank-1 in joint space.
    * **Cross-Modal Generation Fidelity:** Predicts scATAC from scRNA (or vice versa) and computes cell-wise Pearson correlation, RMSE, and cosine similarity.
    * **Direct Peak-to-Gene Linkage:** Correlates promoter/enhancer peak accessibility with target gene transcription across matched single cells.

### 2. Unpaired Multiomics (Independent scRNA-seq and scATAC-seq)
* **Definition:** scRNA-seq and scATAC-seq are profiled from separate single cells derived from the same biological tissue or cell line. Cell barcodes do not match one-to-one.
* **Evaluation in `scSimEval`:**
  * **Unimodal Fidelity:** Evaluates distributions, zero patterns, clustering, batch mixing, marker genes, and trajectories for RNA and ATAC independently.
  * **Cell-Type / Cluster-Level Cross-Modal Transfer (`evaluate_cross_modal_prediction`):** Tests whether a model trained on ATAC accessibility can accurately predict cell types defined in RNA.
  * **Regulatory Modularity (`calc_network_jaccard`, `calc_coexpression_module_fidelity`):** Compares co-accessibility and co-expression network topologies without requiring cell-by-cell pairing.
  * **Multi-Dataset Benchmarking (`evaluate_multiple_datasets`):** Consolidates and ranks simulation accuracy across unpaired modalities side-by-side.

---

# 1. Loading the Packaged Multiomics Dataset

`scSimEval` includes paired scRNA-seq and scATAC-seq matrices (`example_multiomics`):

```{r setup}
library(scSimEval)

# Load packaged multiomics example dataset
data(example_multiomics)

cat("Reference RNA dimensions:", dim(example_multiomics$ref_multi$rna), "\n")
cat("Reference ATAC dimensions:", dim(example_multiomics$ref_multi$atac), "\n")
cat("Simulated RNA dimensions:", dim(example_multiomics$sim_multi$rna), "\n")
cat("Simulated ATAC dimensions:", dim(example_multiomics$sim_multi$atac), "\n")
cat("Cell types:", levels(example_multiomics$cell_types), "\n")
```

The dataset contains:
* `ref_multi`: Named list containing `rna` (count matrix) and `atac` (peak accessibility matrix).
* `sim_multi`: Named list containing simulated `rna` and `atac` matrices.
* `cell_types`: Cell type annotations across the cells.
* `batch_info`: Batch labels across the cells.
* `resource_stats`: Hardware logs (runtime in seconds and peak RAM in MiB).

---

# 2. Category 7: Cross-Modality Coupling Analysis

For multiomics data, `scSimEval` provides specialized functions to evaluate inter-modality coupling.

### 2.1 Alignment in Joint Space (FOSCTTM & Match@1)

Fraction of Samples Closer Than The True Match (FOSCTTM) quantifies how well the two modalities are aligned in joint latent space:

```{r foscttm-eval}
# Compute reduced representations (e.g. PCA) for RNA and ATAC
pca_rna  <- prcomp(t(example_multiomics$ref_multi$rna), rank. = 5)$x
pca_atac <- prcomp(t(example_multiomics$ref_multi$atac), rank. = 5)$x

foscttm_res <- calc_foscttm(x = pca_rna, y = pca_atac)

cat("Bidirectional FOSCTTM score:", round(foscttm_res$foscttm, 4), "\n")
cat("Top-1 Match Rate (Match@1):", round(foscttm_res$match_at_1, 4), "\n")
```

A score of $0.00$ indicates perfect alignment (where each cell's true match is its closest neighbor), while $0.50$ represents random chance.

### 2.2 Cross-Modal Cell Type Label Transfer

We test whether chromatin accessibility profiles in ATAC can accurately predict cell types annotated in RNA:

```{r label-transfer}
transfer_res <- evaluate_cross_modal_prediction(
  mod1_data  = example_multiomics$sim_multi$atac,
  cell_types = example_multiomics$cell_types
)

cat("Cross-Modal Prediction Accuracy:", round(transfer_res$cross_modal_accuracy, 4), "\n")
cat("Cross-Modal Macro F1 Score:", round(transfer_res$cross_modal_F1, 4), "\n")
```

---

# 3. Master Multiomics Pipeline: `evaluate_multiomics_accuracy`

The master function `evaluate_multiomics_accuracy()` coordinates evaluation across all 8 categories:

```{r master-multi}
master_eval <- evaluate_multiomics_accuracy(
  ref_multi  = example_multiomics$ref_multi,
  sim_multi  = example_multiomics$sim_multi,
  cell_types = example_multiomics$cell_types,
  batch_info = example_multiomics$batch_info,
  cpu_time   = example_multiomics$resource_stats$cpu_time,
  memory_mb  = example_multiomics$resource_stats$memory_mb,
  verbose    = FALSE
)

# Inspect tidy master summary table
head(master_eval$benchmark_summary_table)
```

The output contains:
* `benchmark_summary_table`: Master table containing metric names, categories, raw values, and standardized scores.
* `category_scores`: Mean normalized score for each of the 8 categories.
* `composite_score`: Overall unified simulation accuracy index.

---

# 4. Multi-Method / Multi-Dataset Consolidated Benchmarking

To benchmark multiple simulators (or multiple datasets) simultaneously, use `evaluate_multiple_datasets()`:

```{r multi-dataset}
# Create benchmark collection comparing multiple simulated datasets
dataset_collection <- list(
  "Method 1-RNA"  = list(ref = example_multiomics$ref_multi$rna,  sim = example_multiomics$sim_multi$rna),
  "Method 1-ATAC" = list(ref = example_multiomics$ref_multi$atac, sim = example_multiomics$sim_multi$atac),
  "Method 2-RNA"  = list(ref = example_multiomics$ref_multi$rna,  sim = example_multiomics$sim_multi$rna),
  "Method 2-ATAC" = list(ref = example_multiomics$ref_multi$atac, sim = example_multiomics$sim_multi$atac)
)

consolidated <- evaluate_multiple_datasets(
  datasets          = dataset_collection,
  pair_by_prefix    = TRUE,
  compute_bivariate = FALSE,
  verbose           = FALSE
)

# View dataset overview and ranking
print(consolidated$dataset_overview)
```

---

# 5. Visualizing Benchmark Results

`scSimEval` provides built-in functions to generate high-resolution figures:

```r
# 1. Master Benchmark Bubble Matrix across the 8 categories
plot_benchmark_bubble_matrix(
  data  = consolidated,
  title = "Benchmarking Single-Cell Multiomics Simulators"
)

# 2. Evaluation Summary Horizontal Bar Matrix
plot_evaluation_summary(
  data  = consolidated,
  title = "Multiomics Simulator Leaderboard"
)
```

---

# 6. Summary

Whether your study employs paired single-cell multiomics (e.g. 10x Multiome) or independent unpaired datasets, `scSimEval` provides a rigorous, automated, and ground-truth-free framework to evaluate simulation fidelity across all biological and technical dimensions.
